tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-16 paper.tex
Pass through what we do

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from recombination~\citep{wong2023general}, and recent
breakthroughs in inference have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these recent developments, substantial progress has been made
on statistical genetics applications.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate some quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
have been instrumental in making population genetic simulations
of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{kelleher2019,ralph2020,wong2023general},
using the open-source \texttt{tskit} library.

@@

There are therefore sophisticated methods available for simulating
ARGs at the population scale, but there is currently no
easy way to generate quantiative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
used efficient \texttt{tskit} APIs to simulate quantitative
traits for 600,000 humans directly from a simulated ARG.
While the code provided by Martin et al.\ works at this
very large scale, it is tightly coupled to the details of their study
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is therefore a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit}~\citep{ralph2020} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
PyData ecosystem of data-science tools,
using
\texttt{NumPy} arrays~\citep{numpy} and
\texttt{pandas} dataframes~\citep{pandas}
as input and output.

@@

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
Each trait is associated with one or more causal sites (position
on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide). Because we are reasoning
about the inheritance of particular alleles at a given position,
we can consider just the local tree at that site and
the network structure of the ARG along the genome is not
important~\citep{wong2023general}. In
Fig.~\ref{fig:tree-illustration} we show the local tree
relating three diploid individuals for a given causal site.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour). Ancestral
nodes are not associated with individuals.

[WIP]
At a given site, the causal allele is associated with an
effect size $\beta$,
and every node in the tree that carries this allele has a
``genetic value'' of $\beta$.

Following the standard practise in GWAS~\citep{uffelmann2021}
we assume the additive model, and a quantitative trait $y$
is given by
\[
y=X\beta+\epsilon,
\]
where $\beta$ is the effect size, $X$ is a matrix describing the number
of causal alleles in the individual, and $\epsilon$ is the environmental
noise.

Environmental noise is normally distributed with mean zero
and a variance of $V_G(1-h^2)/{h^2}$,
% distribution,
% \begin{align}\label{eq:env}
    % \epsilon\sim \mathcal{N}\left(0,\frac{V_G(1-h^2)}{h^2}\right),
% \end{align}
where $V_G$ is the additive genetic variance from the simulated effect sizes,
and $h^2$ is the narrow-sense heritability provided as input by the user.
The \texttt{tstrait} API is modular, however, and other forms of environmental
noise can be easily generated by the user, if they wish.

\subsection{Effect Sizes}
The first step of a quantitative trait simulation in \texttt{tstrait} is to
define the trait model distribution, as follows:
\begin{lstlisting}[language=Python]
model = tstrait.trait_model(
    distribution='normal', mean=0, var=1)
\end{lstlisting}
\texttt{Tstrait} currently supports effect size simulation from five
commonly used univariate distributions~\citep{gaynor2021,haller2023}
and one multivariate distribution.
The multivariate normal distribution is used to simulate multiple
pleiotropic traits.
The trait model is used as an input of \texttt{sim\_trait} function to simulate
effect sizes, along with the input ARG and number of causal sites:
\begin{lstlisting}[language=Python]
trait_df = tstrait.sim_trait(
    arg, num_causal=5, model=model)
\end{lstlisting}
The output, \texttt{trait\_df} is a \texttt{pandas} dataframe
describing the simulated effect sizes and corresponding
causal site IDs.

\subsection{Genetic Value}
Given a set of causal sites and effect sizes defined in a dataframe
and the input ARG, the \emph{genetic values} [TODO DEFINE ME]
are simulated as follows:
\begin{lstlisting}[language=Python]
genetic_result = tstrait.sim_genetic(
   arg, trait_df, alpha=0)
\end{lstlisting}
The $\alpha$ parameter here is used to
supports effect size simulation from a frequency
dependent model, where the simulated effect sizes from the causal mutation
having allele frequency $p$ are multiplied by $\left(2p(1-p)\right)^\alpha$.
An $\alpha$ value of zero as in this example means there is
no frequency dependence. A negative value of $\alpha$
can be used to model the increased effect sizes on rare variants
resulting from negative selection~\citep{speed2017}.
[TODO is this correct? What's the interpretation of positive alpha?]

@@

Generating the genetic values for a trait is by far the
most computationally intensive part of the simulation,
as we are essentially counting the number
of causal mutations in each individual. This is done by a
straightforward tree traversal algorithm in \texttt{tstrait},
accelerated using the \texttt{numba} JIT compiler for Python.
As we see in later sections, this yields very good performance
at the largest scale, without requiring any additional code in
compiled languages such as C.

\subsection{Phenotype}
The final phenotype simulation is done in \texttt{tstrait}
by imposing some environmental noise.
We assume environmental noise is normally distributed with mean zero
and a variance of
\[
V_G(1-h^2)/{h^2}
\]
where $V_G$ is the additive genetic variance from the simulated effect sizes,
and $h^2$ is the narrow-sense heritability parameter provided as input.
Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
and narrow-sense heritability as parameters:
\begin{lstlisting}[language=Python]
phenotype_df = tstrait.sim_env(
    genetic_result.genetic, h2=0.3)
\end{lstlisting}
This simulation returns a dataframe
describing the simulated genetic value, environmental noise and
resulting phenotype for each individual at each causal site.

As discussed above, there are significant advantages to providing a modular
API. For convenience, we also provide a single
\texttt{sim\_phenotype} function that combines the process of simulating
effect sizes, generating genetic values and imposing environmental noise.

>>>>

Ancestral recombination graphs (ARGs) encode the interwoven paths
of genetic inheritance arising from
% Explanation of citations: people might get offended if we don't cite
% Griffiths when first mentioning the ARG, but Hudson has just as
% much to do with the actual definition that we use in Wong 2023.
recombination~\citep{hudson1983properties,griffiths1997ancestral,wong2023general},
and recent
breakthroughs in inference have opened many new possibilities for their
practical application~\citep{lewanski2023era}.
Building on these recent developments, substantial progress has been made
on statistical genetics applications.
For example,  ARG-based methods
can detect more ultra rare variants than conventional association testing
methods~\citep{zhang2023};
have better power to detect causal loci in
quantitative-trait locus mapping~\citep{link2023tree};
and can provide a sparse and efficient model of linkage disequilibirium
in GWAS and downstream applications~\citep{nowbandegani2023extremely}.

Simulation is a critical component of GWAS method development, and
generally consists of two steps: (1) generate a synthetic genetic
variation dataset; and (2) simulate some quantitative traits
based on this dataset.
(The two steps may sometimes be combined
in a single simulation, or we may use real data instead of simulations in
step (2), but the basic principles are still the same.)
The combined genetic variation and
simulated quantitative traits then represent the ``truth'' against
which association testing methods can be evaluated.
The first step, of generating realistic, large-scale genetic variation
datasets, is a well-studied problem with a mature and robust supporting
software ecosystem. ARG-based methods~\citep{kelleher2016efficient,
kelleher2018efficient,haller2018tree,adrion2020,baumdicker2022efficient,anderson2023}
have been instrumental in making population genetic simulations
of the scale and realism
required to model modern GWAS cohorts~\cite[e.g.][]{martin2017,zaidi2020}.
These simulation methods are all based on the ``succinct tree
sequence'' ARG encoding~\citep{ralph2020,wong2023general},
using the open-source \texttt{tskit} library.

@@

There are therefore sophisticated methods available for simulating
ARGs at the population scale, but there is currently no
easy way to generate quantitative traits based on such an ARG.
Existing methods to simulate quantitative traits from a given set of
genetic sequences in standard formats~\citep[e.g.][]{meyer2018,fernandes2020}
are overwhelmed by the scale of the data. For example,
just parsing the 280TB of VCF for the French-Canadian
dataset mentioned above would take many days of CPU time,
making quantitative trait simulation by this approach unfeasible.
% TODO this needs another pass. We do need to address the Martin et
% al thing up front though, I think.
In their analysis of the portability of polygenic risk
scores across populations, \citet{martin2017}
used efficient \texttt{tskit} APIs to simulate quantitative
traits for 600,000 humans directly from a simulated ARG.
While the code provided by Martin et al.\ works at this
very large scale, it is tightly coupled to the details of their study
and not designed to be reused.
Simulation code can be subtle and difficult to
debug~\citep{ragsdale2020lessons}, and there
is therefore a critical need for a well-documented and
thoroughly tested means of simulating quantitative
traits directly from an ARG.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG
in \texttt{tskit}~\citep{ralph2020} format.
\texttt{Tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
Python data-science ecosytem, allowing users to efficiently analyse
large-scale data using familiar and ergonomic tools.

@@

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
Each trait is associated with one or more causal sites (position
on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide). Because we are reasoning
about the inheritance of alleles at a site on the genome
we can consider just the local tree at that site and
the network structure of the ARG is not important~\citep{wong2023general}.
In Fig.~\ref{fig:tree-illustration} we show the local tree
relating three diploid individuals for a given causal site.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general the ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.

At a given site, the causal allele is associated with an
effect size $\beta$,
and every node in the tree that carries this allele has a
``genetic value'' of $\beta=0.05$. In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele, and therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021}
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is normally distributed with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
% TODO I find this clearer than saying "additive genetic variance"
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG library~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
In the simplest case, simulating a phenotype for an ARG requires
very little code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model, h2=0.3)
\end{lstlisting}
We first create \texttt{model} representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function, which also takes the trait model
and narrow sense heritability $h^2$ as parameters.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{pandas}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conventiently
processed by a host of Python data science tools.

This simple usage is not very flexible, however,
and so \texttt{tstrait} provides more modular building blocks
for power-users, and to potentially integrate with other tools.
Effect sizes can be simulated for a given ARG and trait model
using the \texttt{sim\_trait} function.
This function also supports a form
of frequency dependence, where the a simulated effect size
for a causal allele with frequency $p$
is multiplied by $\left(2p(1-p)\right)^\alpha$.
A negative value of $\alpha$
can be used to model the increased effect sizes on rare variants
resulting from negative selection~\citep{speed2017}.
The return value of the \texttt{sim\_trait} function is a
data frame describing the simulated effect sizes and causal alleles.
This dataframe can then be passed to the \texttt{genetic\_values}
function, which calculates the genetic values for each
node, and accumulates these per individual
(Fig~\ref{fig:tree-illustration}). Finally, the \texttt{sim\_env}
function takes these per-individual genetic values
and adds some simulated environmental noise to produce the final
phenotypes (as discussed above).

@@

% \subsection{Phenotype}
% The final phenotype simulation is done in \texttt{tstrait}
% by imposing some environmental noise.
% We assume environmental noise is normally distributed with mean zero
% and a variance of
% \[
% V_G(1-h^2)/{h^2}
% \]
% where $V_G$ is the additive genetic variance from the simulated effect sizes,
% and $h^2$ is the narrow-sense heritability parameter provided as input.
% Thus, the \texttt{sim\_env} function takes the dataframe of genetic values
% and narrow-sense heritability as parameters:
% \begin{lstlisting}[language=Python]
% phenotype_df = tstrait.sim_env(
%     genetic_result.genetic, h2=0.3)
% \end{lstlisting}
% This simulation returns a dataframe
% describing the simulated genetic value, environmental noise and
% resulting phenotype for each individual at each causal site.

% As discussed above, there are significant advantages to providing a modular
% API. For convenience, we also provide a single
% \texttt{sim\_phenotype} function that combines the process of simulating
% effect sizes, generating genetic values and imposing environmental noise.

\subsection{Implementation}
\texttt{Tstrait} is written entirely in Python. Many numerical operations are
peformed using standard array-oriented approaches and therefore
highly efficient~\citep{numpy}. Generating the genetic values for a given
trait is computationally intensive. This is done by a
straightforward tree traversal algorithm, which is
accelerated using the \texttt{numba} JIT compiler for Python yielding
excellent performance.
